Modelling writing hesitations in text writing as finite mixture process

### 2023-05-29 report/results.Rmd
add alternative spl2 analysis and gunnexp2

```{r include = F}
library(tidyverse)
library(kableExtra)
library(janitor)
library(brms)
library(polspline)
library(patchwork)
library(rmdformats)
library(gridExtra)
library(grid)
library(gtable)
library(gridExtra)
library(scales)
library(ggthemes)
source("../scripts/functions.R")
source("scripts/get_pdens_plot.R")

@@

$$
\begin{align}
\text{iki}_{i} & \sim \theta_\text{location[i]} \cdot \text{LogN}(\beta + \delta_\text{location[i]} + \text{participant}_i, \sigma_{e'_\text{location[i]}}^2) + \\
  & (1 - \theta_\text{location[i]}) \cdot \text{LogN}(\beta + \text{participant}_i, \sigma_{e}^2)\\
\text{where: }
		&\delta \sim \text{N}(0,1)\\
\text{constraint: } & \delta > 0\\
		& \sigma_{e'}^2 > \sigma_{e}^2
\end{align}
$$
This models takes into account two sources of participant-specific error: (1) each participant has an individual fluent typing as in the previous models; (2) each participant has in individual hesitation frequency that differs across levels of the categorical predictor.

@@

# Datasets in overview

Five dataset with keystroke data from text production were used for analysis. An overview can be found in Table \@ref(tab:datasets).

```{r datasets, results='asis'}
tibble(Dataset = c("C2L1", "CATO", "SPL2", "PLanTra", "LIFT"),
                Source = c("@ronneberg2022process",
                              "@torrance2016adolescent",
                              "@torrance",
                              "@rossetti2022s",
                              "@nina_vandermeulen_2020_3893538"),
                Keylogger = c("EyeWrite", 
                           "EyeWrite", 
                           "CyWrite", 
                           "InputLog", 
                           "InputLog"),
                `Writing task` = c(
                                 "Argumentative essays", 
                                 "Expository texts",
                                 "Argumentative essays",
                                 "Text simplification",
                                 "Synthesis"),
                n_ppt = c(126, 26*2, 39, 47, 658),
#                n_texts = c(1, 2, 2, 2, NA),
#                n_sentences = c(),
#                n_words = c(),
                conditions = c(" ", 
                               "weak decoders / control; masked / unmasked",
                               "write in L1 / L2",
                               "pre / post test trained in plain language principles and control", 
                               "Various topics and genres"),
                `Mean age` = c(11.8, 16.9, 20.6, 23, 16.95),
                 Language = c( "Norwegian", 
                               "Norwegian", 
                               "English (L1) / Spanish (L2)", 
                               "English (L2)", 
                               "Dutch")) %>% 
  kable(caption = "Datasets in brief.") %>%
  kable_styling(position = "center", 
                full_width = T, 
                font_size = 12)

@@

Locations with example from Marks

@@

@@

>>>>

```{r include = F}
library(tidyverse)
library(kableExtra)
library(janitor)
library(brms)
library(polspline)
library(patchwork)
library(rmdformats)
library(gridExtra)
library(grid)
library(gtable)
library(scales)
library(ggthemes)
source("../scripts/functions.R")
source("scripts/get_pdens_plot.R")

@@

$$
\begin{align}
\text{iki}_{i} & \sim \theta_\text{location[i]} \cdot \text{LogN}(\beta + \delta_\text{location[i]} + \text{participant}_i, \sigma_{e'_\text{location[i]}}^2) + \\
  & (1 - \theta_\text{location[i]}) \cdot \text{LogN}(\beta + \text{participant}_i, \sigma_{e}^2)\\
\text{where: }
		&\delta \sim \text{N}(0,1)\\
\text{constraint: } & \delta > 0\\
		& \sigma_{e'}^2 > \sigma_{e}^2
\end{align}
$$
This models takes into account two source = s of participant-specific error: (1) each participant has an individual fluent typing as in the previous models; (2) each participant has in individual hesitation frequency that differs across levels of the categorical predictor.

@@

# Datasets

## General overview

Five datasets with keystroke data from text production were used for analysis. An overview can be found in Table \@ref(tab:datasets).

```{r datasets, results='asis'}
tibble(Dataset = c("C2L1", "CATO", "SPL2", "PLanTra", "LIFT"),
                Source = c("@ronneberg2022process",
                              "@torrance2016adolescent",
                              "@torrance",
                              "@rossetti2022s",
                              "@nina_vandermeulen_2020_3893538"),
                Keylogger = c("EyeWrite", 
                           "EyeWrite", 
                           "CyWrite", 
                           "InputLog", 
                           "InputLog"),
                `Writing task` = c(
                                 "Argumentative essays", 
                                 "Expository texts",
                                 "Argumentative essays",
                                 "Text simplification",
                                 "Synthesis"),
                 `N (ppts)` = c(126, 26*2, 39, 47, 658),
#                n_texts = c(1, 2, 2, 2, NA),
#                n_sentences = c(),
#                n_words = c(),
                conditions = c(" ", 
                               "weak decoders / control; masked / unmasked",
                               "write in L1 / L2",
                               "pre / post test trained in plain language principles and control", 
                               "Various topics and genres"),
                `Mean age` = c(11.8, 16.9, 20.6, 23, 16.95),
                 Language = c( "Norwegian", 
                               "Norwegian", 
                               "English (L1) / Spanish (L2)", 
                               "English (L2)", 
                               "Dutch")) %>% 
  kable(caption = "Datasets in brief.") %>%
  kable_styling(position = "center", 
                full_width = T, 
                font_size = 12)

##  Transition types

The transition types that were analysed in this study focuses on those locations that were found, by previous research, to be psycholinguistically meaningful [e.g. @torrance; @chukharev2019combined; @torrance2016adolescent; @de2018exploring] and are detailed in Table \@ref(tab:keyloc). Keytransitions that terminated in an editing operation were excluded from the analysis. Transitions that occurred at the beginning of the text or the beginning of a paragraph were not treated as before-sentence transitions.

```{r keyloc, results='asis'}
tibble(`Transition type` = c("Within word", "Below word", "Before sentence"),
       Description = c("Transitions between any letter",
                       "Keypress after space followed by any letter",
                       "Keypress following a space preceding any letter"),
       Example = c("T^h^e c^a^t m^e^o^w^e^d. T^h^a^t[bsp][bsp]e^n i^t s^l^e^p^t.", 
                   "The ^cat ^meowed. That[bsp][bsp]en ^it ^slept.", 
                   "The cat meowed. ^That[bsp][bsp]en it slept.")) %>% 
  kable(caption = "Transition location classification.") %>%
  footnote("'^' marks transition location, [bsp] represents backspace. IKIs were timed to the shift keypress.") %>% 
  kable_styling(position = "center", 
                full_width = T, 
                font_size = 12)

## Data reduction

```{r reductionfunctions}
# Function to get data reduction infos
get_lift <- function(file, n_samples, n_ppts){
#  file <- "../data/lift.csv"
# n_samples = 50
#  n_ppts = 100
  result <- list()

  d <- read_csv(file)

  full_set <- d %>% 
    count(ppt, topic, genre) %>% 
    summarise(n = n(), .by = ppt) %>% 
    mutate(has_full_set = n == 4)

  result[[3]] <- full_set %>% count(has_full_set)

  ppt_keep <- full_set %>% 
    filter(has_full_set) %>% 
    pull(ppt)

  result[[4]] <- d %>% 
    filter(ppt %in% ppt_keep) %>% 
    select(ppt, enough_sentences) %>% 
    unique() %>% 
    count(enough_sentences)

  result[[1]] <- d %>%
    filter(!is.na(iki), 
           !is_edit,
           ppt %in% ppt_keep,
           enough_sentences) %>%
    select(ppt, iki, location) %>% 
    drop_na() %>% 
    mutate(too_fast = iki <= 50, 
           too_slow = iki >= 30000,
           across(ppt, ~as.numeric(factor(.)))) %>% 
    summarise(across(starts_with("too"), 
                     list(mean = mean, 
                          se = se_bin)),
              .by = c(ppt)) %>% 
    summarise(across(starts_with("too"), 
                     list(mean = mean), 
                     .names = "{.col}"))

  # Sample within each category random data points
  set.seed(365)
  result[[2]] <- d %>% 
    filter(!is.na(iki), 
           !is_edit,
           ppt %in% ppt_keep,
           enough_sentences,
           iki > 50,
           iki < 30000) %>% 
    mutate(keep = 1:n(),
           keep = sample(keep),
           keep = keep <= n_samples,
           .by = c(ppt, location, genre, topic)) %>% 
   summarise(across(keep, 
             list(mean = mean, 
                  se = se_bin)),
              .by = c(ppt, location)) %>% 
   summarise(across(starts_with("keep"), 
                     list(mean = mean), 
                     .names = "{.col}"), 
              .by = location)

   set.seed(365)
   result[[5]] <- d %>% 
    mutate(keep = 1:n(),
           keep = sample(keep),
           keep = keep <= n_samples, 
           .by = c(ppt, location, genre, topic)) %>% 
    filter(keep) %>% 
    mutate(ppt_keep = ppt %in% sample(unique(ppt), n_ppts)) %>% 
    select(ppt, ppt_keep) %>% 
    unique() %>% 
    count(ppt_keep)

   return(result)
}
get_c2l1 <- function(file, n_samples){
  #file <- "../data/c2l1.csv"
  #n_samples <- 100
  d <- read_csv(file) %>% 
    filter(!is.na(iki)) %>% 
    mutate(ppt = as.numeric(factor(subno))) %>% 
    select(ppt, iki, location)

  result <- list()

  result[[1]] <- d %>% 
    mutate(too_fast = iki <= 50,
           too_slow = iki >= 30000) %>%
    summarise(across(starts_with("too"), 
                     list(mean = mean, se = se_bin)),
              .by = c(ppt)) %>% 
    summarise(across(starts_with("too"), 
                     list(mean = mean), 
                     .names = "{.col}"))

  set.seed(365)
  result[[2]] <- d %>% 
    filter(iki > 50,
           iki < 30000) %>% 
    mutate(keep = 1:n(),
           keep = sample(keep),
           keep = keep <= n_samples,
           .by = c(ppt, location)) %>% 
   summarise(across(keep, 
             list(mean = mean, se = se_bin)),
              .by = c(ppt, location)) %>% 
    summarise(across(starts_with("keep"), 
                     list(mean = mean), 
                     .names = "{.col}"), 
              .by = location)

  return(result)
}
get_cato <- function(file, n_samples){
  d <- read_csv(file) %>% 
    filter(!is.na(iki)) %>% 
    mutate(ppt = as.numeric(factor(subno))) %>% 
    select(ppt, iki, location, xn, dystyp)

  result <- list()

  result[[1]] <- d %>% 
    mutate(too_fast = iki <= 50,
           too_slow = iki >= 30000) %>%
    summarise(across(starts_with("too"), 
                     list(mean = mean, 
                          se = se_bin)),
              .by = c(ppt)) %>% 
    summarise(across(starts_with("too"), 
                     list(mean = mean), 
                     .names = "{.col}"))

  set.seed(365)
  result[[2]] <- d %>% 
    filter(iki > 50,
           iki < 30000) %>% 
    mutate(keep = 1:n(),
           keep = sample(keep),
           keep = keep <= n_samples,
           .by = c(ppt, location, xn, dystyp)) %>% 
   summarise(across(keep, 
             list(mean = mean, 
                  se = se_bin)),
              .by = c(ppt, location)) %>% 
   summarise(across(starts_with("keep"), 
                     list(mean = mean), 
                     .names = "{.col}"), 
              .by = location)

  return(result)
}
get_plantra <- function(file, n_samples){
  d <- read_csv(file) %>% 
    filter(!is.na(iki), !is_edit) %>% 
    mutate(ppt = as.numeric(factor(ppt))) %>% 
    select(ppt, iki, location, task)

  result <- list()

  result[[1]] <- d %>% 
    mutate(too_fast = iki <= 50,
           too_slow = iki >= 30000) %>%
    summarise(across(starts_with("too"), 
                     list(mean = mean, 
                          se = se_bin)),
              .by = c(ppt, task)) %>% 
    summarise(across(starts_with("too"), 
                     list(mean = mean), 
                     .names = "{.col}"))

  # keep only participants that did both the pre and post task.
  keep_ppts <- select(d, ppt, task) %>% 
    unique() %>% 
    count(ppt) %>% 
    mutate(both_tasks = n == 2)

  result[[3]] <- count(keep_ppts, both_tasks)

  keep_ppts <- keep_ppts %>% filter(both_tasks) %>% pull(ppt)
  d <- filter(d, ppt %in% keep_ppts)

  # Filter ppts with too few sentences
  keep_ppts <- count(d, ppt, location) %>% 
    filter(location == "before sentence") %>% 
    mutate(enough_sentences = n >= 10)

  result[[4]] <- count(keep_ppts, enough_sentences) %>% 
    rename(more_than_10_sentences = enough_sentences)

  keep_ppts <- keep_ppts %>% 
    filter(enough_sentences) %>% 
    pull(ppt)

  d <- filter(d, ppt %in% keep_ppts)

  set.seed(365)
  result[[2]] <- d %>% 
    filter(iki > 50,
           iki < 30000) %>% 
    mutate(keep = 1:n(),
           keep = sample(keep),
           keep = keep <= n_samples,
           .by = c(ppt, location, task)) %>% 
   summarise(across(keep, 
             list(mean = mean, 
                  se = se_bin)),
              .by = c(ppt, location)) %>% 
   summarise(across(starts_with("keep"), 
                     list(mean = mean), 
                     .names = "{.col}"), 
              .by = location)
  return(result)
}
get_spl2 <- function(file, n_samples){
  result <- list()
#  file <- "../data/spl2.csv"
  d <- read_csv(file) %>% 
    select(-transition_dur) %>% 
    rename(iki = transition_dur_to_mod,
           location = transition_type) %>% 
    filter(!is.na(iki), 
           edit == "noedit") %>% 
    select(ppt = SubNo, iki, location, Lang) %>% 
    drop_na()

  result[[1]] <- d %>% 
    mutate(too_short = iki <= 50,
           too_fast = iki >= 30000,
           across(ppt, ~as.numeric(factor(.)))) %>% 
    summarise(across(starts_with("too"), 
                     list(mean = mean, 
                          se = se_bin)),
              .by = c(ppt)) %>% 
    summarise(across(starts_with("too"), 
                     list(mean = mean), 
                     .names = "{.col}"))

  d <- d %>% filter(iki > 50, iki < 30000)

  # Filter ppts with too few sentences
  keep_ppts <- count(d, ppt, location) %>% 
    filter(location == "sentence_before") %>% 
    mutate(enough_sentences = n >= 10)

  result[[3]] <- count(keep_ppts, enough_sentences) %>% 
    rename(more_than_10_sentences = enough_sentences)

  keep_ppts <- keep_ppts %>% 
    filter(enough_sentences) %>% 
    pull(ppt)

  d <- filter(d, ppt %in% keep_ppts)

  set.seed(365)
  result[[2]] <- d %>% 
    mutate(keep = 1:n(),
           keep = sample(keep),
           keep = keep <= n_samples,
           .by = c(ppt, location, Lang)) %>% 
   summarise(across(keep, 
             list(mean = mean, 
                  se = se_bin)),
              .by = c(ppt, location)) %>% 
   summarise(across(starts_with("keep"), 
                     list(mean = mean), 
                     .names = "{.col}"), 
              .by = location)

  return(result)
}
```

```{r}
# Get info about data reduction
n_samples <- 100
c2l1 <- get_c2l1("../data/c2l1.csv", n_samples = n_samples)
cato <- get_cato("../data/cato.csv", n_samples = n_samples)
lift <- get_lift("../data/lift.csv", n_samples = 50, n_ppts = 100)
plantra <- get_plantra("../data/plantra.csv", n_samples = n_samples)
spl2 <- get_spl2("../data/spl2.csv", n_samples = n_samples)

@@

@@

Add final sample number of ppt, and transitions by location

@@

## GUNNEXP2
